\vspace{-2.0em} Move Schedules: \\ Fast persistence computations in sparse dynamic settings\thanks{This work was partially supported by the National Science Foundation through grants CCF-2006661 and CAREER award DMS-1943758.}\vspace{-0.5em}

### 2021-08-04 move_schedules.tex
Updates from Overleaf
Testing github overleaf merge
Reworded the intro paragraphs to multidimensional persistence

\subsection*{Multidimensional persistence}
A natural question is whether or not there exists a multi-dimensional generalization of persistence. The starting point for this generalization is a structure which filters the space in multiple dimensions simultaneously, a \emph{multifiltration}. Given this structure, the goal of multi-dimensional persistence is to identify persistent features by examining the entire multifiltration. Such a generalization has been thought to have enormous potential in terms of applications~\cite{}. 
For example, one of the often noted drawbacks of persistence is its instability with respect to strong outliers. That is, although persistence diagrams are stable with respect to their input and thus are robust to noise, the presence of outliers can artificially introduce topological features such as loops, obscuring the persistence of significant topological features~\cite{buchet2015topological}. Indeed, this obfuscation can hinder the ability of 1D persistence to uncover ``significant'' topological features, one of the original purposes of persistence. In theory, this formalism enables the practitioner to analyze even noisy data sets directly without resorting to sophisticated data preprocessing techniques which may or may not preserve the underlying topology of the data.

Unfortunately, multi-dimensional persistent homology comes with its own set of new computational and theoretical challenges not necessarily inherited by the 1-D case. The multi-dimensional persistent homology of a multifiltration cannot be captured by complete discrete invariant, and incomplete invariants have proven challenging to use in practice due to the high algorithmic complexity of their corresponding computation. A number of computational strategies have been proposed to counter this, such as homology preserving multi-filtration simplification or parallelization schemes using shared-memory computation~\cite{fugacci2019chunk, lesnick2019computing}. Another strategy is to focus specifically on a fixed dimension, such as 2-D persistence. Utilizing the equivalence between the rank and fibered barcode invariants, Lesnick and Wright~\cite{lesnick2015interactive} developed an elegant way of computing the former via a reparameterization of the latter using standard point-line duality. This clever reparameterization effectively reduces the rank invariant computation to sequence of 1-D barcode computations at ``template points'' lying within the 2-cells of particular planar subdivision of the half-plane $[0, \infty) \times \mathbb{R}$. Indeed, since algorithm~\ref{alg:schedule} was designed for precisely such a computation, the algorithm proposed by~\cite{lesnick2015interactive} is prototypical of the class of methods that stand to benefit from move scheduling.

The purpose of the following sections is to uncover the intrinsic bottlenecks impeding the computation of the rank invariant in practical settings. Towards this goal, we briefly summarize the theory established by several researchers~\cite{lesnick2015interactive, carlsson2010computing, landi2014rank}, focusing on their computational and algorithmic details. 
Our goal is to show that the highest complexity subcomputation of the rank invariant algorithm proposed by~\cite{lesnick2015interactive} is precisely the type of computation we focus on in this effort. 
In doing so, we substantiate our claims from section~\ref{sec:motivation} that our scheduling approach can improve the efficiency of a real-world application of multidimensional persistence by an order of magnitude by comparing Algorithm~\ref{alg:schedule} to several alternatives on a real-world data set and problem: the problem of identifying the underlying topological space of a data set extracting from high-contrast natural images. 
% In what follows, we briefly recount an incomplete but nonetheless very useful invariant of multi-dimensional persistence modules, the \emph{rank invariant}, as well as another invariant which in some sense determines the rank invariant, the \emph{fibered barcode}. It turns out that computing the latter invariant

@@

% Indeed, it was shown in~\cite{lesnick2015interactive} that the fibered barcodes $$ of $M$ (i.e. the collection of barcodes of 1D affine slices of $M$) and the rank invariant determine each other.     
\subsubsection*{Computing the fibered barcode}
%Lesnick and Wright introduced the Rank Invariant Visualization and Exploration Tool (RIVET)~\cite{lesnick2015interactive}, which is both a tool and a collection of computational and mathematical theory which enables exploratory analysis of 2D persistence modules built from arbitrary bifiltrations. From a computational perspective, the core novelty of the RIVET approach is its ability to compute the fibered barcodes efficiently from precomputed information. 
The three dominant subcomputations needed to compute the fibered barcode using the reparameratization discussed above are the following: 
\begin{enumerate}
	\item Computing the bigraded Betti numbers $\xi(M)$ of a given bipersistence module $M$ 
	\item Constructing a line arrangement $\mathcal{A}(M)$ induced by critical lines derived from (1) 
	\item Augmenting $\mathcal{A}(M)$ with a \emph{barcode template} $\mathcal{T}^e$ at every 2-cell $e \in \mathcal{A}(M)$
	%, yielding the \emph{augmented arrangement} $\mathcal{A}^{\sbullet}(M)$ 
\end{enumerate}
It is shown in~\cite{lesnick2019computing} how to compute (1) in approximately $\approx O(m^3)$ using a matrix algorithm similar to Algorithm~\ref{alg:reduce}. Constructing and storing the line arrangement $\mathcal{A}(M)$ with $n$ lines and $k$ vertices can be achieved using known algorithms in computational geometry (e.g. the Benteley-Ottmann algorithm) in output-sensitive $O((n+k) \log n)$ time and $O(\kappa^2)$ memory, where $\kappa$ is a coarseness parameter associated with the computation of $\xi(M)$. 
Finally, the computation of the barcodes templates $\mathcal{T}$ (step 3) takes on the order of $O(m^3 \kappa + m \kappa^2 \log \kappa)$ elementary operations and $O(m \kappa^2)$ storage.

>>>>

\subsection*{Multidimensional persistence}
A natural question is whether or not there exists a multidimensional generalization of persistence. Given a structure which filters the space in multiple dimensions simultaneously, a \emph{multifiltration}, the goal of multidimensional persistence is to identify persistent features by examining the entire multifiltration. Such a generalization has appeared naturally in many application contexts, showing potential as a tool for exploratory data analysis~\cite{lesnick2012multidimensional}. 
For example, one of the often noted drawbacks of persistence is its instability with respect to strong outliers. Although persistence diagrams are stable with respect to their input and thus are robust to noise, the presence of strong outliers can artificially modify the connectivity of the underlying filtration, obscuring the detection of what would otherwise be significant topological structures~\cite{buchet2015topological}. Indeed, since inferring the homology of some underlying topological space around which data are distributed was one of the original purposes of persistence\footnote{One of the first uses of persistence was to infer the topology of attractors in dynamical systems from experimental data\cite{perea2018brief}.}, strong outliers pose a significant barrier to the widespread adoption of persistent homology. By filtering the space according to e.g. some notion of density, multidimensional persistence enables the practitioner to handle outliers directly, without resorting to sophisticated data preprocessing techniques which may or may not preserve the underlying topology of the data.

Unfortunately, multidimensional persistent homology comes with its own set of computational and theoretical challenges not necessarily inherited by the 1-D case. There is not complete discrete invariant for multidimensional persistence, and the computations associated to other incomplete invariants exhibit high algorithmic complexity. A number of computational strategies have been proposed to counter this, such as homology preserving multifiltration simplification or parallelization schemes using shared-memory computation~\cite{fugacci2019chunk}. Rather than attempting to tackle multiple dimensions simultaneously, another approach is to focus on a fixed dimension, such as 2-D persistence. Utilizing the equivalence between the rank and fibered barcode invariants, Lesnick and Wright~\cite{lesnick2015interactive} developed an elegant way of visualizing the former via a reparameterization of the latter using standard point-line duality. This clever reparameterization effectively reduces the fibered barcode computation to a sequence of 1-D barcode computations at ``template points'' lying within the 2-cells of particular planar subdivision of the half-plane $[0, \infty) \times \mathbb{R}$. Indeed, since algorithm~\ref{alg:schedule} was designed for precisely such a computation, the algorithm proposed by~\cite{lesnick2015interactive} is prototypical of the class of methods that stand to benefit from move scheduling.

The purpose of the following sections is to uncover the intrinsic bottlenecks impeding the computation of the fibered barcode invariant in practical settings. Towards this goal, we briefly summarize the theory established by several researchers~\cite{lesnick2015interactive, carlsson2010computing, landi2014rank}, focusing on their computational and algorithmic details. 
Our goal is to show that the highest complexity subcomputation of the fibered barcode invariant algorithm proposed by~\cite{lesnick2015interactive} is precisely the type of computation we focus on in this effort. 
In doing so, we substantiate our claims from section~\ref{sec:motivation} that our scheduling approach can improve the efficiency of a real-world application of multidimensional persistence by an order of magnitude by comparing Algorithm~\ref{alg:schedule} to several alternatives on a real-world data set and problem: the problem of identifying the underlying topological space of a data set extracting from high-contrast natural images. 
% In what follows, we briefly recount an incomplete but nonetheless very useful invariant of multi-dimensional persistence modules, the \emph{rank invariant}, as well as another invariant which in some sense determines the rank invariant, the \emph{fibered barcode}. It turns out that computing the latter invariant

@@

% Indeed, it was shown in~\cite{lesnick2015interactive} that the fibered barcodes $$ of $M$ (i.e. the collection of barcodes of 1D affine slices of $M$) and the rank invariant determine each other.     
\subsubsection*{Computing the fibered barcode}
%Lesnick and Wright introduced the Rank Invariant Visualization and Exploration Tool (RIVET)~\cite{lesnick2015interactive}, which is both a tool and a collection of computational and mathematical theory which enables exploratory analysis of 2D persistence modules built from arbitrary bifiltrations. From a computational perspective, the core novelty of the RIVET approach is its ability to compute the fibered barcodes efficiently from precomputed information. 
The three dominant subcomputations needed to compute the fibered barcode using the reparameratization discussed above are the following: 
\begin{enumerate}
	\item Computing the bigraded Betti numbers $\xi(M)$ of a given bipersistence module $M$ 
	\item Constructing a line arrangement $\mathcal{A}(M)$ induced by critical lines derived from (1) 
	\item Augmenting $\mathcal{A}(M)$ with a \emph{barcode template} $\mathcal{T}^e$ at every 2-cell $e \in \mathcal{A}(M)$
	%, yielding the \emph{augmented arrangement} $\mathcal{A}^{\sbullet}(M)$ 
\end{enumerate}
It is shown in~\cite{lesnick2019computing} how to compute (1) in approximately $\approx O(m^3)$ using a matrix algorithm similar to Algorithm~\ref{alg:reduce}. Constructing and storing the line arrangement $\mathcal{A}(M)$ with $n$ lines and $k$ vertices is related to the \emph{line segment intersection problem}, which known algorithms in computational geometry can solve in output-sensitive $O((n+k) \log n)$ time (see~\cite{boissonnat2000robust} for an overview). The number of 2-cells in $\mathcal{A}(M)$ and $O(\kappa^2)$ memory, where $\kappa$ is a coarseness parameter associated with the computation of $\xi(M)$. 
Finally, the computation of the barcodes templates $\mathcal{T}$ (step 3) takes on the order of $O(m^3 \kappa + m \kappa^2 \log \kappa)$ elementary operations and $O(m \kappa^2)$ storage.
